---
title: "The report of Writeup project"
author: "Yang Li"
date: "1/9/2026"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

## Introduction
The objective is to develop a predictive model to forecast the manner in which an exercise was performed, identified by the "class" variable in the training dataset.
The report should detail:

1. The methodology used to build the model.

2. The application of cross-validation techniques.

3. The estimation of the expected out-of-sample error.

4. A justification for all modeling decisions made.

5. The use of the final model to predict 20 distinct test cases.

## Data exploration and preparation
The following preparation of the data are done.
0. Load data
```{r echo=TRUE}
library(caret)
library(dplyr)
library(ggplot2)
library(lattice)

training_orig <- read.csv("pml-training.csv")
predicting_orig <- read.csv("pml-testing.csv")
```

1. Remove the features, which the procentage of their values "NA" exceed threshold.
```{r echo=TRUE}
is_majority_na <- function(col, threshold = 0.7) {
  nr_prop = sum(is.na(col)) / length(col)
  if (nr_prop > threshold) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

get_cols_to_remove <- function(dataframe, threshold = 0.7) {
  result_bool <- sapply(dataframe, is_majority_na, threshold = threshold)
  cols <- names(result_bool[result_bool == TRUE])
  return(cols)
}

cols_to_remove <- get_cols_to_remove(training_orig, threshold = 0.7)

training1 <- training_orig %>%
  select(-all_of(cols_to_remove))

predicting1 <- predicting_orig %>%
  select(-all_of(cols_to_remove))

```
2. Remove the features, which the procentage of their values "" exceed threshold.
```{r echo=TRUE}
get_cols_with_high_empty_string_prop <- function(dataframe, threshold = 0.7) {
  empty_prop <- colMeans(dataframe == "")
  result_bool <- empty_prop > threshold
  cols_to_remove <- names(result_bool[result_bool == TRUE])
  return(cols_to_remove)
}

cols_to_remove <- get_cols_with_high_empty_string_prop(training1, threshold = 0.7)

training2 <- training1 %>%
  select(-all_of(cols_to_remove))

predicting2 <- predicting1 %>%
  select(-all_of(cols_to_remove))
```

3. manuall remove features
```{r echo=TRUE}
manual_to_remove <- c("user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "X", "new_window", "num_window")
training3 <- training2 %>%
  select(-all_of(manual_to_remove))
predicting3 <- predicting2 %>%
  select(-all_of(manual_to_remove))
```

4. For some missing values of the filtered features, I will assign the average value from the same feature

```{r echo=TRUE}
impute_with_mean <- function(col) {
  avg_value <- mean(col, na.rm = TRUE)
  
  col[is.na(col)] <- avg_value
  
  return(col)
}
training <- training3 %>%
  mutate_if(is.numeric, impute_with_mean)
predicting <- predicting3 %>%
  mutate_if(is.numeric, impute_with_mean)

training$classe <- as.factor(training$classe)
predicting$problem_id <- as.factor(predicting$problem_id)

```

3. Divide the training set to training and testing set
```{r echo=TRUE}
inTrain = createDataPartition(training$classe, p=0.7, , list=FALSE)
training = training[ inTrain,]
testing = training[-inTrain,]
```

## Model building
I choose decision tree based approach because of the following reasons:

1. Classification Task: The target variable (classe) is categorical (A, B, C, D, E). The decision tree method is a classification algorithm designed to predict these categories accurately. Multivariate regression is used for continuous, numerical outcomes.

2. High Dimensionality: The dataset has hundreds of features. The decision tree method, such as Random Forest, handles high-dimensional data robustly, naturally dealing with feature selection and overfitting. 

3. Performance: In practice, Random Forest consistently achieves higher accuracy (often near 100%) on this specific dataset compared to regression-based approaches.

I prepare cross validation. I use standard k-folder cross validation, where k is 5 and it repeats only once because of restriction of running time.
```{r echo=TRUE}
set.seed(1000)
control_params <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 1
)
```

I use the "Recursive Partitioning and Regression Trees" (rpart) method to train the data
```{r echo=TRUE}
modFit_rpart <- train(
  classe ~ ., 
  data = training, 
  method = "rpart",      
  trControl = control_params
)
print(modFit_rpart)
res_rpart <- predict(modFit_rpart, newdata=testing)
conf_matrix <- confusionMatrix(res_rpart, testing$classe)
res <- conf_matrix$overall['Accuracy']
cat("\nAccuracy:\n")
cat(res, "\n")
cat("\nOut of sample error:\n")
cat(1 - res, "\n")
```

I use the "Randam Forest" (rf) method to train the data
```{r echo=TRUE}
modFit_rf <- train(
  classe ~ ., 
  data = training, 
  method = "rf",      
  trControl = control_params
)
print(modFit_rf)
res_rf <- predict(modFit_rf, newdata=testing)
conf_matrix <- confusionMatrix(res_rf, testing$classe)
res <- conf_matrix$overall['Accuracy']
cat("\nAccuracy:\n")
cat(res, "\n")
cat("\nOut of sample error:\n")
cat(1 - res, "\n")
```

I use the "Gradient Boosting Machine" (gbm) method to train the data
```{r echo=TRUE}
modFit_gbm <- train(
  classe ~ ., 
  data = training,
  method = "gbm",     
  trControl = control_params,
  verbose = FALSE 
)
print(modFit_gbm)
res_gbm <- predict(modFit_gbm, newdata=testing)
conf_matrix <- confusionMatrix(res_gbm, testing$classe)
res <- conf_matrix$overall['Accuracy']
cat("\nAccuracy:\n")
cat(res, "\n")
cat("\nOut of sample error:\n")
cat(1 - res, "\n")
```
According to the comparison of above 3 models, one can see that random forest method deliver the best results, following by the "gbm". The "rpart" method gives the worst results.

## Prediction:
I use the "random forest" method to predict the new test cases in "pml-training.cvs".
```{r echo=TRUE}
res <- predict(modFit_rf, newdata=predicting)
print("\nPredicted results:\n")
print(res)
```

