---
title: "Baysian linear regression with dataset swiss"
author: "Yang Li"
format: pdf
editor: visual
---

## Understand the problem

The `swiss` dataset is a classic benchmarking dataset built into R. It contains socioeconomic indicators and fertility rates for **47 French-speaking provinces** (districts) of Switzerland around the year **1888**. At this time, Switzerland was undergoing a massive demographic transition.

The primary objective when working with the `swiss` dataset is **to model and predict the Fertility Rate** of a province based on its socioeconomic characteristics.

## Plan and properly collect relevant data

The following graph shows us a overlook of the data:

```{r}
head(swiss)
```

We can see all data type are integer and we can build a linear regression model to the data.

```{r}
#| echo: false
str(swiss)
```

Because most algorithms (e.g., linear regression or neural networks) cannot perform calculations with empty cells and R usually aborts the training process with an error message if `NA` values are present in the dataset, we need to check if there are any NULL values in the dataset and fix them accordingly

```{r}
any(is.na(df))
```

FALSE means no NULL values. \## Explore data The next step is to check the correlation between variables. Optically, we can see the variables Agriculture, Examination and Education are correlated.

```{r}
pairs(swiss)
```

This is confirmed with the correlation matrix, but the absolute values in the correlation matrix are still below the 0.7, so we will build 3 models with including the correlated variable and excluding the correlated models and compare the quality of them.

```{r}
cor_matrix <- cor(swiss[, -1])
round(cor_matrix, 3)
```

## Postulate, Fit and Check Models

### Model 1

A model built with all variables

```{r}
#| fig-width: 9
#| fig-height: 4

library("rjags")
mod_string1 = " model {
    for (i in 1:length(Fertility)) {
        Fertility[i] ~ dnorm(mu[i], prec)
        mu[i] = b0 + b[1]*Agriculture[i] + b[2]*Examination[i] + b[3]*Education[i] + b[4]*Catholic[i] + b[5]*Infant.Mortality[i]
        res[i] <- Fertility[i] - mu[i]
    }
    
    b0 ~ dnorm(0.0, 1.0/1.0e6)
    for (i in 1:5) {
        b[i] ~ dnorm(0.0, 1.0/1.0e6)
    }
    
    prec ~ dgamma(1.0/2.0, 1.0*1500.0/2.0)
    sig2 = 1.0 / prec
    sig = sqrt(sig2)
} "

data_jags = list(
  Fertility = swiss$Fertility, 
  Agriculture = swiss$Agriculture,
  Examination = swiss$Examination,
  Education = swiss$Education,
  Catholic = swiss$Catholic,
  Infant.Mortality = swiss$Infant.Mortality
)

set.seed(72)

params = c("b0", "b", "sig", "res")

inits = function() {
  inits = list("b0" = rnorm(1, 0.0, 100.0),
               "b"=rnorm(5, 0.0, 100.0), 
               "prec"=rgamma(1, 1.0, 1.0))
}

mod1 = jags.model(textConnection(mod_string1), data=data_jags, inits=inits, n.chains=3)
update(mod1, 1000) # burn-in

mod1_sim = coda.samples(model=mod1,
                        variable.names=params,
                        n.iter=10000)

mod1_csim = do.call(rbind, mod1_sim)
plot(mod1_sim[, c("b0", "b[1]")])
plot(mod1_sim[, c("b[2]", "b[3]")])
plot(mod1_sim[, c("b[4]", "b[5]")])
plot(mod1_sim[, c("sig")])
```

In the following show the auto correlation graph, residual graph and Deviance Information Criterion (DIC). Auto correlation plot:

```{r}
autocorr.plot(mod1_sim)

```

Residual Plot

```{r}
autocorr.plot(mod1_sim)
res_cols <- grep("res", colnames(mod1_csim), value = TRUE)
mean_residual <- colMeans(mod1_csim[, res_cols])
plot(mean_residual, 
     main = "Residualplot (Bayesian)", 
     ylab = "Residual", 
     xlab = "Index", 
     pch = 19, col = "darkblue")
abline(h = 0, col = "red", lwd = 2)
```

DIC

```{r}
dic_result1 = dic.samples(mod1, n.iter=10000)
print(dic_result1)

```

### Model 2

A model built without examination

```{r}
#| fig-width: 9
#| fig-height: 4

library("rjags")
mod_string2 = " model {
    for (i in 1:length(Fertility)) {
        Fertility[i] ~ dnorm(mu[i], prec)
        mu[i] = b0 + b[1]*Agriculture[i] + b[2]*Education[i] + b[3]*Catholic[i] + b[4]*Infant.Mortality[i]
        res[i] <- Fertility[i] - mu[i]
    }
    
    b0 ~ dnorm(0.0, 1.0/1.0e6)
    for (i in 1:4) {
        b[i] ~ dnorm(0.0, 1.0/1.0e6)
    }
    
    prec ~ dgamma(1.0/2.0, 1.0*1500.0/2.0)
    sig2 = 1.0 / prec
    sig = sqrt(sig2)
} "

data_jags = list(
  Fertility = swiss$Fertility, 
  Agriculture = swiss$Agriculture,
  Education = swiss$Education,
  Catholic = swiss$Catholic,
  Infant.Mortality = swiss$Infant.Mortality
)

set.seed(72)

inits = function() {
  inits = list("b0" = rnorm(1, 0.0, 100.0),
               "b"=rnorm(4, 0.0, 100.0), 
               "prec"=rgamma(1, 1.0, 1.0))
}

mod2 = jags.model(textConnection(mod_string2), data=data_jags, inits=inits, n.chains=3)
update(mod2, 1000) # burn-in

mod2_sim = coda.samples(model=mod2,
                        variable.names=params,
                        n.iter=10000)

mod2_csim = do.call(rbind, mod2_sim)
plot(mod2_sim[, c("b0", "b[1]")])
plot(mod2_sim[, c("b[2]", "b[3]")])
plot(mod2_sim[, c("b[4]", "sig")])
```

In the following show the auto correlation graph, residual graph and Deviance Information Criterion (DIC). Auto correlation plot:

```{r}
autocorr.plot(mod2_sim)

```

Residual Plot

```{r}
autocorr.plot(mod2_sim)
res_cols <- grep("res", colnames(mod2_csim), value = TRUE)
mean_residual <- colMeans(mod2_csim[, res_cols])
plot(mean_residual, 
     main = "Residualplot (Bayesian)", 
     ylab = "Residual", 
     xlab = "Index", 
     pch = 19, col = "darkblue")
abline(h = 0, col = "red", lwd = 2)
```

DIC

```{r}
dic_result2 = dic.samples(mod2, n.iter=10000)
print(dic_result2)

```
