---
title: "GoogleTrend"
author: "Yang Li"
date: "2026-06-01"
format: pdf
output:
  pdf_document: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Understanding the problem

### Objective

The goal of this analysis is to model and understand the underlying structural dynamics of global Google Trends search interest for the keyword "time series" from 2004 to the present.

### Key Challenges

Non-Stationarity: The raw search volume data exhibits a strong, long-term global trend, making it non-stationary and unsuitable for standard stationary time series models.

Seasonality-Effecte: The data contains a distinct annual (12-month) seasonal pattern that needs to be isolated

Unknown Variance: Real-world search data experiences changing fluctuations over time (heteroscedasticity). Therefore, the model must account for an unknown observation variance to ensure reliable and robust interval estimation.

Methodology:To address these challenges, a Normal Dynamic Linear Model (NDLM) with unknown variance is constructed. The model decomposes the raw time series directly by combining a polynomial trend component (local level) with a trigonometric seasonal component ($q=1$ harmonic). An baysian based filter and smoother are applied, utilizing an optimized discount factor to dynamically track structural changes.

## Data exploration

```{r}
timeseries_data <- read.csv("googletrends.csv", skip = 2, header = TRUE)
colnames(timeseries_data)[1] <- "date"
colnames(timeseries_data)[2] <- "hits"
timeseries_data$date = as.Date(timeseries_data$date)
Y_raw <- as.numeric(timeseries_data$hits)
acf(Y_raw, main = "ACF for das Google-Trend 'time series' Raw data")
Y <- diff(Y_raw)
acf(Y, main = "ACF for the transformed Google-Trend 'time series'")
```

The graph shows Rapid Decay Towards Zero in contrast to the first image, the bars here drop immediately and abruptly after lag 0. Most lags are now very close to zero and well within the blue dashed lines. This proves that the long-term dependency (the global trend) has completely disappeared.

Although the series is stationary, two bars clearly stand out beyond the blue significance line: Lag 12: Shows a distinct positive correlation (approx. +0.3). Lag 24: Also shows a positive correlation (approx. +0.26). Meaning: Since these are monthly Google Trends data, lag 12 represents exactly 12 months (1 year). This means that the time series has a strong annual seasonality. The search behavior for "time series" repeats in a yearly pattern (e.g., always at the start of university semesters).

## Construct NDLM

Since the series is now trend-stationary but shows clear seasonality at lag 12, it is perfectly suited for a polynomial Trends with seasonal componenten.

In addition to a constant mean, a constant variance over time is the second mandatory requirement. If the variance changes over time (heteroscedasticity), predictions become unreliable, therefore we need to give an estimate about variance in the model (unknown variance).

When using the dlm (Dynamic Linear Models) package in R, which works directly with the raw data. The package does not automatically difference the data, so we need to include a one order linear trend component.

```{r, results = 'hide'}
data=list(yt=timeseries_data$hits)

library(dlm)
model_seasonal=dlmModTrig(s=12,q=1,dV=0,dW=1)
model_trend=dlmModPoly(order=1,dV=10,dW=rep(1,1), m0=c(40))
model=model_trend+model_seasonal
model$C0=10*diag(3)
n0=1
S0=10
h = 5
k=length(model$m0)
T=length(data$yt)

Ft=array(0,c(1,k,T))
Gt=array(0,c(k,k,T))
for(t in 1:T){
  Ft[,,t]=model$FF
  Gt[,,t]=model$GG
}

source('all_dlm_functions_unknown_v.r')
source('discountfactor_selection_functions.r')

matrices=set_up_dlm_matrices_unknown_v(Ft=Ft,Gt=Gt)
initial_states=set_up_initial_states_unknown_v(model$m0,
                                               model$C0,n0,S0)

df_range=seq(0.7,1,by=0.01)
print(length(data$yt))
print(T)
print(h)
results_MSE <- adaptive_dlm(data, matrices, 
                            initial_states, df_range,"MSE",forecast=FALSE)

print(paste("The selected discount factor:",results_MSE$df_opt))
```

## Filtering and Smoothing

```{r}
results_filtered <- results_MSE$results_filtered
ci_filtered <- get_credible_interval_unknown_v(ft = results_filtered$ft, Qt = results_filtered$Qt, nt =   results_filtered$nt
)
```

```{r}
results_smoothed <- results_MSE$results_smoothed
ci_smoothed <- get_credible_interval_unknown_v(results_smoothed$fnt, results_smoothed$Qnt,  results_filtered$nt[length(results_smoothed$fnt)]
)
```

### Plot of linear trend components

This script configures the graphical layout and generates two separate time series plots to analyze Google Trends data alongside smoothed trend estimates.

```{r}
index = timeseries_data$date
par(mfrow=c(1,1))
plot(index, data$yt, pch=19, cex=0.3, col='lightgray', 
     xlab="time", ylab="Google hits", main="Trend Level & Distributions")
lines(index, results_smoothed$mnt[,1], lwd=2, col='magenta')
lines(index, results_filtered$mt[,1], lwd=1.5, col='blue')
lines(index, ci_filtered[, 1], lty=3, col='blue', lwd=1)
lines(index, ci_filtered[, 2], lty=3, col='blue', lwd=1)
par(mfrow=c(1,1))
plot(index, results_smoothed$mnt[,2], col='darkblue', lwd=2, type='l', 
     ylim=c(-0.6, 0.6), xlab="time", ylab="rate of change", main="Slope Component")
abline(h=0, col='red', lty=2)
par(mfrow=c(1,1))
```
### Plot of seasonal components

This code generates a line plot to visualize a seasonal or cyclical component—specifically with a period of 12 (likely representing monthly patterns)—extracted from the smoothed time series results.

```{r}
plot(index, results_smoothed$mnt[, 3], type = "l", lwd = 2, col = "darkgreen", ylim = c(-12, 12), xlab = "time", 
     ylab = "", main = "period=12"
)
```

## Forcast

Once the model is filtered up to the time threshold $T-h$ using the optimal discount factor, the forecast_function_unknown_v function projects the system into the future for $h = 5$ steps:

State Projection: It projects the state parameters (at) and their uncertainties (Rt) iteratively using the transition matrix $G_t$. If a discount factor is present, the uncertainty matrix Rt is scaled up by $1/\delta$ at each step to reflect growing future uncertainty.

Observation Projection: It projects these states into the observation space to produce point forecasts (ft) and forecast variances (Qt).

Distribution Properties: Because the true observation variance is unknown and estimated sequentially ($S_t$), the resulting forecast distribution follows a Student-$t$ distribution rather than a standard Gaussian distribution, providing wider, safer confidence intervals for the predictions.
```{r}
posterior_states <- list(
  mt = results_filtered$mt[1:(T - h), , drop = FALSE],
  Ct = results_filtered$Ct[, , 1:(T - h), drop = FALSE],
  St = results_filtered$St[1:(T - h)]
)

results_forecast <- forecast_function_unknown_v(posterior_states = posterior_states, k = h, matrices = matrices,        
  delta = results_MSE$df_opt)

forecast_sd <- sqrt(results_forecast$Qt)
lower_bound <- results_forecast$ft - 2 * forecast_sd
upper_bound <- results_forecast$ft + 2 * forecast_sd
y_limits    <- range(c(lower_bound, upper_bound))

plot(1:5, results_forecast$ft, type = "b", pch = 19, col = "blue",  ylim = y_limits, xaxt = "n",
     main = "95% Forecast Distribution Bounds (h = 5)", 
     xlab = "Future Time Step (k)", ylab = "Predicted Value")

axis(1, at = 1:5)
lines(1:5, upper_bound, lty = 2, col = "red", lwd = 1.5)
lines(1:5, lower_bound, lty = 2, col = "red", lwd = 1.5)

polygon(c(1:5, 5:1), c(upper_bound, rev(lower_bound)), col = rgb(1, 0, 0, 0.1), border = NA)

legend("topleft", legend = c("Point Forecast (Distribution Mean)", "95% Confidence Space (Distribution Width)"), 
       col = c("blue", "red"), lty = c(1, 2), pch = c(19, NA), bty = "n")
```  
### Summary of Results

The analysis successfully specified and evaluated an adaptive Bayesian Dynamic Linear Model for the "time series" Google Trends dataset.

### Key Findings

Model Optimization: The grid search identified 0.7 as the optimal discount factor ($\delta_{opt} = 0.7$), indicating that the model benefits from a relatively high adaptation rate to capture recent structural shifts.

Trend Extraction: The model successfully decoupled the underlying long-term growth from noise, capturing the massive spike in search interest observed after 2025.

Seasonal Decomposition: The trigonometric seasonal component ($q=1$) effectively isolated the steady 12-month cyclical behavior (annual peaks and troughs) hidden within the raw data.

Uncertainty Quantification: By implementing a DLM with unknown variance, the model provided robust 95% credible intervals that accurately reflect the estimation uncertainty over the entire timeline.

### Final Takeaway:

The structural decomposition demonstrates that interest in "time series" is not only growing exponentially in recent years but is heavily driven by predictable annual cycles. This adaptive state-space framework provides a solid foundation for generating reliable future out-of-sample forecasts.
